---
title: "CCHS example"
author: "Juan Li and the recodeflow contributors"
date: 2025-07-02
format:
  html:
    toc: true
    html-math-method: katex
    css: styles.css
editor: visual
---

::: callout-note
## About this vignette

All numeric values shown in this vignette are computed from the actual CCHS sample metadata files. Code is hidden by default for readability, but you can view the source `.qmd` file to see how values are calculated.
:::

## Overview

This example demonstrates generating mock Canadian Community Health Survey (CCHS) data using cchsflow metadata. The generated mock data can be used to test harmonization workflows before accessing real CCHS data.

```{r setup}
#| message: false
library(MockData)
library(dplyr)
library(stringr)
```

## Load metadata

Load the CCHS harmonization metadata files that define variables and their details across cycles.

```{r load-metadata}
# CCHS sample variables
variables <- read.csv(
  system.file("extdata/cchs/variables_cchsflow_sample.csv", package = "MockData"),
  header = TRUE,
  check.names = FALSE,
  na.strings = c("", "NA", "N/A"),
  stringsAsFactors = FALSE
)

# CCHS sample variable details
variable_details <- read.csv(
  system.file("extdata/cchs/variable_details_cchsflow_sample.csv", package = "MockData"),
  header = TRUE,
  check.names = FALSE,
  na.strings = c("", "NA", "N/A"),
  stringsAsFactors = FALSE
)
```

The CCHS sample metadata includes `r nrow(variables)` harmonized variables with `r nrow(variable_details)` detail rows.

## Extract available cycles

```{r extract-cycles}
#| echo: false
# Extract unique cycle names from databaseStart column
cycles <- sort(unique(unlist(str_split(paste(variables$databaseStart, collapse = ","), ","))))
cycles <- str_trim(cycles[str_detect(cycles, "cchs")])
```

The sample metadata includes `r length(cycles)` CCHS cycles: `r paste(cycles, collapse = ", ")`.

## Understanding the metadata

```{r examine-cycle}
#| echo: false
# Pick the first cycle to explore
example_cycle <- cycles[1]

# Get all harmonized variables available in this cycle
cycle_vars <- get_cycle_variables(example_cycle, variables, variable_details)
```

The sample metadata includes 20 harmonized variables. For `r example_cycle`, there are `r nrow(cycle_vars)` harmonized variables available:

```{r show-cycle-vars}
#| echo: false
knitr::kable(head(cycle_vars[, c("variable", "variable_raw", "variableType", "label")], 10))
```

## Get raw variables to generate

For mock data generation, we need to create the **raw source variables** (before harmonization), not the harmonized variables.

```{r get-raw-variables}
# Get unique raw variables needed for this cycle (excludes derived variables)
raw_vars <- get_raw_variables(example_cycle, variables, variable_details,
                               include_derived = FALSE)
```

```{r compute-raw-stats}
#| echo: false
# Separate by type for easier processing
raw_cat <- raw_vars[raw_vars$variableType == "Categorical", ]
raw_con <- raw_vars[raw_vars$variableType == "Continuous", ]
```

This cycle requires `r nrow(raw_vars)` unique raw variables: `r nrow(raw_cat)` categorical and `r nrow(raw_con)` continuous.

```{r show-raw-vars}
#| echo: false
knitr::kable(raw_vars)
```

## Generate mock data for one cycle

Now let's generate mock data for a single cycle.

```{r configure-generation}
# Configuration
n_records <- 100
target_cycle <- example_cycle
seed <- 12345

# Initialize data frame
df_mock <- data.frame(id = 1:n_records)
```

We'll generate `r n_records` mock records for `r target_cycle`.

### Generate categorical variables

```{r generate-categorical}
#| echo: false
# Get raw categorical variables for this cycle
raw_cat_vars <- get_raw_variables(target_cycle, variables, variable_details,
                                   include_derived = FALSE)
raw_cat_vars <- raw_cat_vars[raw_cat_vars$variableType == "Categorical", ]

for (i in 1:nrow(raw_cat_vars)) {
  var_raw <- raw_cat_vars$variable_raw[i]

  result <- create_cat_var(
    var_raw = var_raw,
    cycle = target_cycle,
    variable_details = variable_details,
    variables = variables,
    length = n_records,
    df_mock = df_mock,
    seed = seed + i
  )

  if (!is.null(result)) {
    df_mock <- cbind(df_mock, result)
  }
}
```

Generated `r nrow(raw_cat_vars)` categorical variables. Data frame now has `r ncol(df_mock)` columns.

### Generate continuous variables

```{r generate-continuous}
#| echo: false
# Get raw continuous variables for this cycle
raw_con_vars <- get_raw_variables(target_cycle, variables, variable_details,
                                   include_derived = FALSE)
raw_con_vars <- raw_con_vars[raw_con_vars$variableType == "Continuous", ]

for (i in 1:nrow(raw_con_vars)) {
  var_raw <- raw_con_vars$variable_raw[i]

  result <- create_con_var(
    var_raw = var_raw,
    cycle = target_cycle,
    variable_details = variable_details,
    variables = variables,
    length = n_records,
    df_mock = df_mock,
    distribution = "uniform",
    seed = seed + 100 + i
  )

  if (!is.null(result)) {
    df_mock <- cbind(df_mock, result)
  }
}
```

Generated `r nrow(raw_con_vars)` continuous variables. Final data frame has `r ncol(df_mock)` columns.

### Examine the result

**Mock data structure:**

```{r examine-result}
#| echo: false
str(df_mock)
```

**First 5 rows:**

```{r show-first-rows}
#| echo: false
knitr::kable(head(df_mock, 5))
```

```{r check-missing}
#| echo: false
na_counts <- colSums(is.na(df_mock))
has_missing <- any(na_counts > 0)
```

**Missing values:** `r if(has_missing) paste(sum(na_counts > 0), "variables have missing values") else "No missing values"`

## Summary

This example demonstrated generating mock CCHS data for testing cchsflow harmonization workflows. The generated data:

-   Respects category ranges from variable_details
-   Includes appropriate missing values
-   Uses reproducible seeds
-   Can be used to test harmonization functions before accessing real CCHS data

## Next steps

-   Test your cchsflow harmonization pipeline on this mock data
-   Generate mock data for additional cycles as needed
-   Calculate derived variables after harmonization (not during mock data generation)