---
title: "Example of generating mock data using CHMS files based on variable and variable-details sheets"
author: "Juan Li"
date: 2025-10-21
format:
  html:
    toc: true
    html-math-method: katex
editor: visual
---

## Overview

The Canadian Health Measures Survey (CHMS) data exists only in secure data environments. This example demonstrates how to generate mock CHMS data for testing chmsflow workflows.

## Setup

```{r}
#| warning: false
#| message: false
#| output: false

library(dplyr)       # Data manipulation
library(stringr)     # Working with strings

# Load MockData package functions
# Note: When running as vignette, use library(MockData)
# For development, source directly:
source("../R/mockdata-parsers.R")
source("../R/mockdata-helpers.R")
source("../R/create_cat_var.R")
source("../R/create_con_var.R")

```

## Read in variable and variable_details sheets

```{r}
# CHMS sample variables
variables <- read.csv(system.file("extdata/chms/chmsflow_sample_variables.csv", package = "MockData"),
                     header = T, check.names=FALSE, na.strings=c("","NA","N/A"))
# CHMS sample variable details
variable_details <- read.csv(system.file("extdata/chms/chmsflow_sample_variable_details.csv", package = "MockData"),
                            header = T, check.names=FALSE, na.strings=c("","NA","N/A"))
```

## Get required cycles

```{r}
# Extract unique cycle names from databaseStart column
cycles <- sort(unique(unlist(str_split(paste(variables$databaseStart, collapse = ","), ","))))
cycles <- str_trim(cycles[str_detect(cycles, "cycle")])
print("Available CHMS cycles:")
print(cycles)
```

## Understanding the metadata

The sample metadata includes 20 harmonized variables. Let's examine how they're structured for one cycle:

```{r}
# Pick the first cycle to explore
example_cycle <- cycles[1]
cat("Examining cycle:", example_cycle, "\n\n")

# Get all harmonized variables available in this cycle
cycle_vars <- get_cycle_variables(example_cycle, variables, variable_details)

cat("Total harmonized variables in", example_cycle, ":", nrow(cycle_vars), "\n")
cat("Sample of variables:\n")
print(head(cycle_vars[, c("variable", "variable_raw", "variableType", "label")], 10))
```

## Get raw variables to generate

For mock data generation, we need to create the **raw source variables** (before harmonization), not the harmonized variables:

```{r}
# Get unique raw variables needed for this cycle (excludes derived variables)
raw_vars <- get_raw_variables(example_cycle, variables, variable_details,
                               include_derived = FALSE)

cat("\nUnique raw variables to generate:", nrow(raw_vars), "\n")
print(raw_vars)

# Separate by type for easier processing
raw_cat <- raw_vars[raw_vars$variableType == "Categorical", ]
raw_con <- raw_vars[raw_vars$variableType == "Continuous", ]

cat("\n- Categorical:", nrow(raw_cat), "variables")
cat("\n- Continuous:", nrow(raw_con), "variables\n")
```

## Generate mock data for one cycle

Now let's generate mock data for a single cycle using the modern helper-driven approach:

```{r}
# Configuration
n_records <- 100
target_cycle <- example_cycle
seed <- 12345

# Initialize data frame
df_mock <- data.frame(id = 1:n_records)

cat("Generating", n_records, "mock records for", target_cycle, "\n\n")
```

### Generate categorical variables

```{r}
# Get raw categorical variables for this cycle
raw_cat_vars <- get_raw_variables(target_cycle, variables, variable_details,
                                   include_derived = FALSE)
raw_cat_vars <- raw_cat_vars[raw_cat_vars$variableType == "Categorical", ]

cat("Generating", nrow(raw_cat_vars), "categorical variables:\n")
for (i in 1:nrow(raw_cat_vars)) {
  var_raw <- raw_cat_vars$variable_raw[i]
  cat("  -", var_raw, "\n")

  result <- create_cat_var(
    var_raw = var_raw,
    cycle = target_cycle,
    variable_details = variable_details,
    variables = variables,
    length = n_records,
    df_mock = df_mock,
    seed = seed + i
  )

  if (!is.null(result)) {
    df_mock <- cbind(df_mock, result)
  }
}

cat("\nCategorical variables generated. Data frame now has", ncol(df_mock), "columns\n")
```

### Generate continuous variables

```{r}
# Get raw continuous variables for this cycle
raw_con_vars <- get_raw_variables(target_cycle, variables, variable_details,
                                   include_derived = FALSE)
raw_con_vars <- raw_con_vars[raw_con_vars$variableType == "Continuous", ]

cat("Generating", nrow(raw_con_vars), "continuous variables:\n")
for (i in 1:nrow(raw_con_vars)) {
  var_raw <- raw_con_vars$variable_raw[i]
  cat("  -", var_raw, "\n")

  result <- create_con_var(
    var_raw = var_raw,
    cycle = target_cycle,
    variable_details = variable_details,
    variables = variables,
    length = n_records,
    df_mock = df_mock,
    distribution = "uniform",
    seed = seed + 100 + i
  )

  if (!is.null(result)) {
    df_mock <- cbind(df_mock, result)
  }
}

cat("\nContinuous variables generated. Data frame now has", ncol(df_mock), "columns\n")
```

### Examine the result

```{r}
# Show structure
cat("\nMock data structure:\n")
str(df_mock)

# Show first few rows
cat("\nFirst 10 rows:\n")
print(head(df_mock, 10))

# Check for missing values
cat("\nMissing value summary:\n")
na_counts <- colSums(is.na(df_mock))
print(na_counts[na_counts > 0])
```

## Summary

This example demonstrates generating mock CHMS data for testing chmsflow workflows. The generated data:

- Respects category ranges from variable_details
- Includes tagged NA values
- Uses reproducible seeds
- Can be used for testing harmonisation functions

For production use, generate all variables across all cycles using the approach shown in `mockdata-tools/test-all-cycles.R`.
