---
title: "Choose a MockData v0.4 backend"
format: html
vignette: >
  %\VignetteIndexEntry{Choose a MockData v0.4 backend}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| label: setup
#| include: false
input_file <- tryCatch(knitr::current_input(dir = TRUE), error = function(e) NULL)
candidate_roots <- unique(c(
  ".",
  "..",
  if (!is.null(input_file)) file.path(dirname(input_file), "..")
))
package_root <- NULL
for (candidate in candidate_roots) {
  description <- file.path(candidate, "DESCRIPTION")
  if (file.exists(description) &&
      any(grepl("^Package:\\s+MockData\\s*$", readLines(description, warn = FALSE)))) {
    package_root <- candidate
    break
  }
}

if (!is.null(package_root)) {
  devtools::load_all(package_root, quiet = TRUE)
} else {
  library(MockData)
}
```

::: {.vignette-about}
**About this vignette:** This how-to explains when to use the default native
backend and when to try the optional `simstudy` backend. The `simstudy` examples
run when `simstudy >= 0.8.1` is installed and otherwise render a clear message.
:::

## The short version

Use the native backend by default.

```{r}
spec <- mock_spec(
  mock_spec_continuous("age", range = c(18, 85), rtype = "integer"),
  mock_spec_categorical(
    "smoking",
    levels = c("never", "former", "current"),
    proportions = c(0.5, 0.3, 0.2),
    rtype = "character"
  )
)

native_data <- generate_mock_data_native(spec, n = 100, seed = 101)
head(native_data)
```

The native backend is always available, stays within MockData's MIT-licensed
code, and is the backend used by `create_mock_data()` for supported v0.4
metadata.

Use the optional `simstudy` backend when you want to exercise that engine path
or when future MockData features need simulation mechanics that `simstudy`
already provides.

## Check whether simstudy is available

MockData keeps `simstudy` optional. It is listed in `Suggests`, not `Imports`,
so installing MockData does not require installing `simstudy`.

```{r}
simstudy_available <- requireNamespace("simstudy", quietly = TRUE) &&
  utils::packageVersion("simstudy") >= numeric_version("0.8.1")

simstudy_available
```

If `simstudy` is unavailable, use `generate_mock_data_native()`.

```{r}
if (!simstudy_available) {
  message(
    "The optional simstudy backend is not available in this R environment; ",
    "using generate_mock_data_native() is the recommended path."
  )
}
```

## Run the same spec through both backends

For categorical variables and uniform continuous variables, both backends can
generate the baseline data.

```{r}
native_large <- generate_mock_data_native(spec, n = 2000, seed = 202)

if (simstudy_available) {
  simstudy_large <- generate_mock_data_simstudy(spec, n = 2000, seed = 202)
  head(simstudy_large)
} else {
  simstudy_large <- NULL
}
```

When `simstudy` is installed, compare broad properties rather than expecting
row-for-row equality. The engines use different internals.

```{r}
if (simstudy_available) {
  c(
    native_mean_age = mean(native_large$age),
    simstudy_mean_age = mean(simstudy_large$age)
  )
}
```

```{r}
if (simstudy_available) {
  rbind(
    native = prop.table(table(factor(
      native_large$smoking,
      levels = c("never", "former", "current")
    ))),
    simstudy = prop.table(table(factor(
      simstudy_large$smoking,
      levels = c("never", "former", "current")
    )))
  )
}
```

## Mixed specs are allowed

The optional backend uses `simstudy` only for pieces it can currently generate
safely. Other variables route through MockData's native backend inside the same
call.

```{r}
mixed_spec <- mock_spec(
  mock_spec_categorical(
    "smoking",
    levels = c("never", "former", "current"),
    proportions = c(0.5, 0.3, 0.2),
    rtype = "character"
  ),
  mock_spec_continuous(
    "bmi",
    range = c(15, 50),
    distribution = "normal",
    mean = 27,
    sd = 5,
    rtype = "double"
  ),
  mock_spec_date(
    "interview_date",
    range = as.Date(c("2020-01-01", "2020-12-31"))
  )
)

mixed_native <- generate_mock_data_native(mixed_spec, n = 100, seed = 303)
head(mixed_native)
```

```{r}
if (simstudy_available) {
  mixed_simstudy <- generate_mock_data_simstudy(mixed_spec, n = 100, seed = 303)
  head(mixed_simstudy)
}
```

In this example, `smoking` can be generated through `simstudy`; `bmi` and
`interview_date` stay native because MockData owns the truncated normal and
calendar-date contracts in v0.4.

## Post-processing stays MockData-owned

Missing codes, garbage values, and diagnostics are applied after baseline
generation. That is true for both backends.

```{r}
post_spec <- mock_categorical(
  "response",
  levels = c("1", "97"),
  proportions = c(0.6, 0.4),
  rtype = "character",
  missing_codes = "97",
  missing_proportions = 0.2,
  garbage_rules = list(low = list(proportion = 0.1, range = "[-2, 0]"))
)

native_baseline <- generate_mock_data_native(post_spec, n = 100, seed = 404)
native_processed <- postprocess_mock_data(native_baseline, post_spec, seed = 405)

names(attr(native_processed, "mockdata_diagnostics")$variables$response)
```

```{r}
if (simstudy_available) {
  simstudy_baseline <- generate_mock_data_simstudy(post_spec, n = 100, seed = 404)
  simstudy_processed <- postprocess_mock_data(simstudy_baseline, post_spec, seed = 405)

  names(attr(simstudy_processed, "mockdata_diagnostics")$variables$response)
}
```

The diagnostics shape is the same because post-processing is not delegated to
`simstudy`.

## License and dependency posture

MockData is MIT licensed. `simstudy` is GPL-3 licensed. Keeping `simstudy`
optional lets MockData keep the core package MIT while still allowing users to
try the advanced backend when that dependency is acceptable in their project.

If your workflow needs no optional dependency, use:

```{r}
generate_mock_data_native(spec, n = 10, seed = 1)
```

If your workflow explicitly wants to test the optional backend and `simstudy` is
installed, use:

```{r}
if (simstudy_available) {
  generate_mock_data_simstudy(spec, n = 10, seed = 1)
}
```

## Decision guide

Choose the native backend when:

- you want the default v0.4 behavior;
- you need MockData to work without optional dependencies;
- you are generating categorical, continuous, date, missing-code, or garbage
  examples covered by the native pipeline;
- you want the simplest path for package tests and vignettes.

Try the optional `simstudy` backend when:

- `simstudy >= 0.8.1` is already acceptable in your project;
- you want to exercise the optional engine path;
- you are preparing for future features where `simstudy` provides mature
  simulation mechanics;
- you still want MockData to own missing-code, garbage-value, and diagnostics
  semantics after generation.
