---
title: "MockData v0.4 design philosophy"
format: html
vignette: >
  %\VignetteIndexEntry{MockData v0.4 design philosophy}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| label: setup
#| include: false
input_file <- tryCatch(knitr::current_input(dir = TRUE), error = function(e) NULL)
candidate_roots <- unique(c(
  ".",
  "..",
  if (!is.null(input_file)) file.path(dirname(input_file), "..")
))
package_root <- NULL
for (candidate in candidate_roots) {
  description <- file.path(candidate, "DESCRIPTION")
  if (file.exists(description) &&
      any(grepl("^Package:\\s+MockData\\s*$", readLines(description, warn = FALSE)))) {
    package_root <- candidate
    break
  }
}

if (!is.null(package_root)) {
  devtools::load_all(package_root, quiet = TRUE)
} else {
  library(MockData)
}
```

::: {.vignette-about}
**About this vignette:** This explanation describes why MockData v0.4 is shaped
around `mock_spec`, native generation, optional `simstudy`, and MockData-owned
post-processing. It is not a tutorial; start with the v0.4 getting-started
vignette if you want a first workflow.
:::

## Mock data, not synthetic data

MockData generates mock data for package development, QA, documentation,
examples, and training. Its output is meant to exercise code paths. It is not
intended for privacy release, inference, or population-valid statistical
analysis.

That boundary is deliberate. In health-data and survey-data settings,
"synthetic data" can imply privacy review, data-sharing obligations, or
statistical validity claims. MockData avoids that claim. It helps you test a
pipeline before you have access to real data; it does not replace real data for
analysis.

The working sentence is:

> Give a recodeflow-style specification a body, so you can test the recoding
> before you have the data.

## The people v0.4 is trying to serve

Three user groups shaped the v0.4 design.

First, recodeflow ecosystem maintainers need data frames that can run through
cchsflow, chmsflow, and recodeflow examples and tests. They already have
`variables.csv` and `variable_details.csv`, so MockData should read those
metadata files rather than invent a competing file format.

Second, methodologists and package authors need examples and vignettes that run
without restricted data. They may want a small, readable direct API rather than
a full metadata table.

Third, QA developers need deliberately bad data. Out-of-range ages, declared
missing codes, impossible dates, and invalid category values are not incidental
features; they are the point when testing validation code.

v0.4 tries to serve all three without making any one workflow the only workflow.

## Why `mock_spec` exists

Before v0.4, MockData's generators read metadata, parsed ranges, generated
values, applied missing codes, injected garbage, coerced types, and assembled
columns in one path. That was useful while the package was young, but it made
validation, backend choice, and diagnostics hard to reason about.

v0.4 introduces `mock_spec` as the normalized internal representation. Different
front doors can produce the same spec:

- direct helpers, such as `mock_continuous()` and `mock_categorical()`
- composable constructors, such as `mock_spec_continuous()`
- recodeflow metadata through `mock_spec_from_recodeflow()`

The spec is then consumed by generation and post-processing layers.

```{r}
spec <- mock_spec(
  mock_spec_continuous(
    "age",
    range = c(18, 85),
    distribution = "normal",
    mean = 50,
    sd = 12,
    rtype = "integer"
  ),
  mock_spec_categorical(
    "smoking",
    levels = c("never", "former", "current"),
    proportions = c(0.5, 0.3, 0.2),
    rtype = "character"
  )
)

names(spec$variables)
```

This is the main architectural move: parse once, validate once, then generate
from the normalized shape.

## Two tiers, one model

The direct helpers are there for the first ten minutes.

```{r}
one_variable <- mock_continuous(
  "age",
  range = c(18, 85),
  distribution = "normal",
  mean = 50,
  sd = 12,
  rtype = "integer"
)

names(one_variable$variables)
```

The lower-level constructors are there when you want to compose multiple
variables or build adapters.

```{r}
same_variable <- mock_spec(
  mock_spec_continuous(
    "age",
    range = c(18, 85),
    distribution = "normal",
    mean = 50,
    sd = 12,
    rtype = "integer"
  )
)

names(same_variable$variables)
```

Those are two surface syntaxes for the same internal model. That is why the
package can support small hand-written examples and recodeflow metadata without
duplicating generation logic.

## Why the backend is hybrid

MockData v0.4 has a native backend and an optional `simstudy` backend.

The native backend is the default. It is always available, keeps MockData usable
without optional dependencies, and owns the simple cases that are central to the
package: categorical values, continuous values, dates, missing-code semantics,
garbage values, and diagnostics.

`simstudy` is optional. It is a mature GPL-3 simulation package with useful
machinery for future advanced features, but MockData remains MIT licensed by
keeping `simstudy` in `Suggests` and soft-gating the backend.

```{r}
native_data <- generate_mock_data_native(spec, n = 5, seed = 1)
native_data
```

```{r}
simstudy_available <- requireNamespace("simstudy", quietly = TRUE) &&
  utils::packageVersion("simstudy") >= numeric_version("0.8.1")

if (simstudy_available) {
  generate_mock_data_simstudy(spec, n = 5, seed = 1)
} else {
  message("simstudy is not installed; the native backend remains available.")
}
```

This split is intentionally conservative. MockData should not reimplement a
large simulation library when a good one exists, but it also should not make a
GPL-3 package mandatory for users who only need the core mock-data path.

## Why post-processing is separate

Missing codes and garbage values are not just another distribution. They are QA
semantics layered on top of otherwise valid generated data.

v0.4 therefore generates baseline values first, then applies missing-code and
garbage rules in a separate post-processing pass.

```{r}
qa_spec <- mock_categorical(
  "response",
  levels = c("1", "97"),
  proportions = c(0.7, 0.3),
  rtype = "character",
  missing_codes = "97",
  missing_proportions = 0.2
)

baseline <- generate_mock_data_native(qa_spec, n = 100, seed = 11)
processed <- postprocess_mock_data(baseline, qa_spec, seed = 12)

diagnostics <- attr(processed, "mockdata_diagnostics")
names(diagnostics$variables$response)
```

The diagnostics matter because a value can naturally collide with a declared
missing code. In the example above, `97` is both a valid level and a missing
code. MockData records which rows naturally drew `97` and which rows were
assigned `97` during post-processing.

```{r}
response_diag <- diagnostics$variables$response

c(
  preexisting = length(response_diag$preexisting_missing_code_indices),
  assigned = length(response_diag$assigned_missing_indices)
)
```

That distinction is what makes the output auditable for QA workflows.

## Why strictness increased

Earlier MockData versions were often permissive: warn, skip a variable, and
return whatever could be generated. That behavior was convenient in exploratory
work, but risky in package tests and documentation. A silently missing column
can make a vignette or downstream test look successful while testing the wrong
thing.

v0.4 moves toward strict generation for the new pipeline. Unsupported features
should either fail loudly or route through an explicit compatibility path.

`create_mock_data()` keeps compatibility by retaining legacy fallback routes,
especially for `validate = FALSE`, `variable_details = NULL`, detail-level
`databaseStart` filtering, and unsupported native-backend features. Use
`verbose = TRUE` while migrating so the chosen path is visible.

## What is deliberately deferred

Several features are intentionally not solved in v0.4.

Formula-derived variables are detected and kept loud rather than silently
ignored. They need a dependency-aware evaluator, sandboxing rules, and clear
syntax.

Multi-variable correlation and richer joint distributions are future work.
`simstudy` is one possible engine for those features, but v0.4 does not claim to
generate statistically realistic joint distributions.

Table 1 bootstrap is also future work. It is a natural third adapter: take
published descriptive statistics and produce a `mock_spec`. That is useful, but
it should not be squeezed into the recodeflow adapter.

LinkML or another schema-first model remains a possible north star for the
larger recodeflow ecosystem. v0.4 keeps the internal spec abstract enough that a
future schema adapter could produce it.

These are roadmap items, not hidden guarantees.

## How the v0.4 refactor was reviewed

The v0.4 architecture was developed through a spike, milestone PRs, and repeated
review of code, tests, silent-failure paths, and documentation. That process
changed the design in concrete ways:

- strict-by-default behavior became more important than permissive fallback;
- diagnostics became a first-class auditability contract;
- `simstudy` stayed optional to preserve MockData's dependency and license
  posture;
- a Phase C communication note made sibling-package testing part of the release
  process;
- executable vignettes became part of validation, not just prose.

The development notes in `development/` and maintainer-only review notes
preserve more of that review trail. This vignette distills the user-facing
design choices.

## The design in one paragraph

MockData v0.4 normalizes inputs into `mock_spec`, validates that shape strictly
by default, generates baseline values through a native backend or optional
`simstudy` backend, and then applies MockData-owned post-processing for missing
codes, garbage values, and diagnostics. It keeps recodeflow metadata central,
adds simpler direct APIs, and preserves the public `create_mock_data()` wrapper
for compatibility. It is mock data for development and QA, not synthetic data
for inference.
