---
title: "Inspect diagnostics and garbage rules in MockData v0.4"
format: html
vignette: >
  %\VignetteIndexEntry{Inspect diagnostics and garbage rules in MockData v0.4}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| label: setup
#| include: false
input_file <- tryCatch(knitr::current_input(dir = TRUE), error = function(e) NULL)
candidate_roots <- unique(c(
  ".",
  "..",
  if (!is.null(input_file)) file.path(dirname(input_file), "..")
))
package_root <- NULL
for (candidate in candidate_roots) {
  description <- file.path(candidate, "DESCRIPTION")
  if (file.exists(description) &&
      any(grepl("^Package:\\s+MockData\\s*$", readLines(description, warn = FALSE)))) {
    package_root <- candidate
    break
  }
}

if (!is.null(package_root)) {
  devtools::load_all(package_root, quiet = TRUE)
} else {
  library(MockData)
}
```

::: {.vignette-about}
**About this vignette:** This how-to shows how to inspect the
`mockdata_diagnostics` attribute added by the v0.4 post-processing layer. The
examples focus on audit trails for missing-code collisions and garbage-value
rules.
:::

## Why diagnostics matter

Mock data often needs two kinds of unusual values:

- missing codes, such as `97` or `999`
- garbage values, such as impossible ages used to test validation code

Sometimes a value can be both meaningful and suspicious. For example, code `97`
could be a valid category in one source file and also a declared missing code in
another. MockData records diagnostics so you can tell whether a value was drawn
naturally by the baseline generator or assigned later by post-processing.

## Create a collision case

Start with a categorical variable where `97` is both a valid level and a
declared missing code. This is deliberately awkward; it is the case diagnostics
are designed to make auditable.

```{r}
response_spec <- mock_categorical(
  "response",
  levels = c("1", "97"),
  proportions = c(0.65, 0.35),
  rtype = "character",
  missing_codes = "97",
  missing_proportions = 0.20
)

baseline <- generate_mock_data_native(response_spec, n = 200, seed = 11)
table(baseline$response)
```

The baseline already contains some `97` values because `97` is a valid level.
Now apply post-processing.

```{r}
processed <- postprocess_mock_data(baseline, response_spec, seed = 12)
table(processed$response)
```

## Read the diagnostics

Diagnostics live in a data-frame attribute.

```{r}
diagnostics <- attr(processed, "mockdata_diagnostics")
names(diagnostics$variables)
```

For a variable, two fields are especially important:

- `preexisting_missing_code_indices`: rows whose baseline value already matched
  a declared missing code
- `assigned_missing_indices`: rows changed by post-processing to a missing code

```{r}
response_diag <- diagnostics$variables$response

length(response_diag$preexisting_missing_code_indices)
length(response_diag$assigned_missing_indices)
```

These two sets should be distinct.

```{r}
intersect(
  response_diag$preexisting_missing_code_indices,
  response_diag$assigned_missing_indices
)
```

Both groups contain `97` in the final data, but they mean different things.

```{r}
head(processed$response[response_diag$preexisting_missing_code_indices])
head(processed$response[response_diag$assigned_missing_indices])
```

Use the diagnostics when your tests need to distinguish a naturally drawn
collision from a missing code assigned by MockData.

## Add garbage rules

Garbage rules deliberately inject invalid or out-of-range values. Here `age`
has one missing code and two garbage rules:

- `low`: values below the valid age range
- `high`: values above the valid age range

```{r}
age_spec <- mock_continuous(
  "age",
  range = c(18, 85),
  distribution = "normal",
  mean = 50,
  sd = 12,
  rtype = "integer",
  missing_codes = 999,
  missing_proportions = 0.05,
  garbage_rules = list(
    high = list(proportion = 0.03, range = "[120, 150]"),
    low = list(proportion = 0.04, range = "[0, 17]")
  )
)

age_baseline <- generate_mock_data_native(age_spec, n = 200, seed = 21)
age_processed <- postprocess_mock_data(age_baseline, age_spec, seed = 22)
```

MockData applies garbage rules in canonical order: `low`, then `high`, then any
other named rules in caller order. The diagnostics use the same order.

```{r}
age_diag <- attr(age_processed, "mockdata_diagnostics")$variables$age
names(age_diag$assigned_garbage_indices)
```

Inspect the assigned rows.

```{r}
low_idx <- age_diag$assigned_garbage_indices$low
high_idx <- age_diag$assigned_garbage_indices$high

length(low_idx)
range(age_processed$age[low_idx])

length(high_idx)
range(age_processed$age[high_idx])
```

Missing-code rows are protected from garbage assignment.

```{r}
intersect(age_diag$assigned_missing_indices, low_idx)
intersect(age_diag$assigned_missing_indices, high_idx)
```

## Combine variables in one pipeline

Most workflows generate several variables together. The same diagnostics shape
is used for every variable in the spec.

```{r}
spec <- mock_spec(
  response_spec$variables$response,
  age_spec$variables$age
)

combined_baseline <- generate_mock_data_native(spec, n = 200, seed = 31)
combined_processed <- postprocess_mock_data(combined_baseline, spec, seed = 32)

combined_diag <- attr(combined_processed, "mockdata_diagnostics")
names(combined_diag$variables)
```

A compact audit summary can be built from the diagnostics.

```{r}
data.frame(
  variable = names(combined_diag$variables),
  preexisting_missing = vapply(
    combined_diag$variables,
    function(x) length(x$preexisting_missing_code_indices),
    integer(1)
  ),
  assigned_missing = vapply(
    combined_diag$variables,
    function(x) length(x$assigned_missing_indices),
    integer(1)
  ),
  assigned_garbage = vapply(
    combined_diag$variables,
    function(x) sum(lengths(x$assigned_garbage_indices)),
    integer(1)
  )
)
```

## Preserve diagnostics before reshaping

Diagnostics are stored as an attribute on the returned data frame. Some
downstream operations keep attributes and others drop them. If diagnostics are
part of your QA workflow, save them before heavy reshaping or joins.

```{r}
saved_diagnostics <- attr(combined_processed, "mockdata_diagnostics")

subset_data <- combined_processed[1:5, ]
is.null(attr(subset_data, "mockdata_diagnostics"))

names(saved_diagnostics$variables)
```

## Re-running post-processing

`postprocess_mock_data()` is intentionally not idempotent. Running it again on a
data frame that already has `mockdata_diagnostics` would double-contaminate the
data, so MockData stops loudly.

```{r}
#| error: true
postprocess_mock_data(combined_processed, spec, seed = 33)
```

Start again from baseline data when you want a fresh post-processing draw.
