---
title: "MockData for recodeflow users"
format: html
vignette: >
  %\VignetteIndexEntry{MockData for recodeflow users}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| label: setup
#| include: false
# Load package - works in both local dev and pkgdown build
if (file.exists("../DESCRIPTION")) {
  devtools::load_all("../", quiet = TRUE)
} else {
  library(MockData)
}
```

::: {.vignette-about}
**About this vignette:** This tutorial is for users who already have recodeflow metadata files (`variables.csv` and `variable_details.csv`). If you're new to MockData, start with [Getting started](getting-started.html) instead.
:::

## Quick start

If you already have recodeflow metadata files, you're one function call away from generating mock data:

```{r}
#| eval: false

library(MockData)

# Point MockData at your existing metadata files
mock_data <- create_mock_data(
  databaseStart = "cchs2001",  # Your database identifier
  variables = "path/to/variables.csv",
  variable_details = "path/to/variable_details.csv",
  n = 1000,
  seed = 123
)

head(mock_data)
```

That's it. MockData reads your existing metadata and generates mock data that matches your variable specifications.

## What MockData reads from your files

MockData uses the same metadata files as cchsflow and chmsflow. No duplicate specifications needed.

### From variables.csv

MockData reads:

- `variable`: Variable name in the generated dataset
- `rType`: R data type (factor, character, integer, double, date)
- `role`: Filters for enabled variables only
- `distribution`: Distribution type (uniform, normal, exponential, gompertz) for continuous/date variables
- `mean`, `sd`, `rate`, `shape`: Distribution parameters
- `event_prop`, `followup_min`, `followup_max`: Survival data parameters
- `prop_garbage`, `garbage_low_prop`, `garbage_high_prop`: Data quality testing parameters

**Example:**

```{r}
#| echo: false

# Load minimal-example metadata
variables <- read.csv(
  system.file("extdata/minimal-example/variables.csv", package = "MockData"),
  stringsAsFactors = FALSE,
  check.names = FALSE
)

knitr::kable(head(variables[, c("variable", "rType", "role", "distribution")], 5))
```

### From variable_details.csv

MockData reads:

- `recStart`: Category codes or value ranges using interval notation (e.g., `[18,100]` for continuous, `[2001-01-01,2005-12-31]` for dates)
- `recEnd`: Classification (`copy`, `NA::a`, `NA::b`)
- `catLabel`: Category labels
- `proportion`: Category proportions (optional)

**Example:**

```{r}
#| echo: false

# Load minimal-example metadata
details <- read.csv(
  system.file("extdata/minimal-example/variable_details.csv", package = "MockData"),
  stringsAsFactors = FALSE,
  check.names = FALSE
)

knitr::kable(head(details[details$variable == "smoking", c("variable", "recStart", "recEnd", "catLabel", "proportion")], 4))
```

## The databaseStart parameter

The `databaseStart` parameter tells MockData which database/cycle to generate data for. This is the same identifier you use in recodeflow workflows.

```{r}
#| eval: false

# Generate data for CCHS 2001
mock_cchs2001 <- create_mock_data(
  databaseStart = "cchs2001_p",  # Match your database identifier
  variables = "variables.csv",
  variable_details = "variable_details.csv",
  n = 1000
)

# Generate data for CHMS Cycle 1
mock_chms1 <- create_mock_data(
  databaseStart = "cycle1",  # Match your database identifier
  variables = "variables.csv",
  variable_details = "variable_details.csv",
  n = 1000
)
```

MockData filters `variable_details.csv` to only generate variables where the `databaseStart` column matches your specified database.

## Working example with minimal-example metadata

Let's generate mock data using the minimal-example metadata included with MockData:

```{r}
#| warning: false
#| message: false

# Load recodeflow-compatible metadata
variables <- read.csv(
  system.file("extdata/minimal-example/variables.csv", package = "MockData"),
  stringsAsFactors = FALSE,
  check.names = FALSE
)

variable_details <- read.csv(
  system.file("extdata/minimal-example/variable_details.csv", package = "MockData"),
  stringsAsFactors = FALSE,
  check.names = FALSE
)

# Generate mock data
mock_data <- create_mock_data(
  databaseStart = "minimal-example",
  variables = variables,
  variable_details = variable_details,
  n = 100,
  seed = 456
)

# View structure
cat("Generated", nrow(mock_data), "observations across", ncol(mock_data), "variables\n\n")
```

**View sample data:**

```{r}
#| echo: false

head(mock_data[, 1:5])
```

## Common workflows

### Testing harmonisation code

Use mock data to test cchsflow or chmsflow harmonisation before accessing real data:

```{r}
#| eval: false

library(cchsflow)

# 1. Generate mock raw data
mock_raw <- create_mock_data(
  databaseStart = "cchs2001_p",
  variables = "variables.csv",
  variable_details = "variable_details.csv",
  n = 1000
)

# 2. Apply harmonisation
mock_harmonised <- rec_with_table(
  data = mock_raw,
  variables = variables,
  variable_details = variable_details,
  databaseStart = "cchs2001_p"
)

# 3. Test your analysis code
mock_harmonised %>%
  filter(age_der >= 65) %>%
  summarise(prevalence_bmi_obesity = mean(bmi_der_cat == "Obese", na.rm = TRUE))
```

### Developing analysis scripts

Write and debug analysis scripts before data access:

```{r}
#| eval: false
#| warning: false
#| message: false

library(dplyr)

# Generate mock data
mock_data <- create_mock_data(
  databaseStart = "cchs2001_p",
  variables = "variables.csv",
  variable_details = "variable_details.csv",
  n = 5000,
  seed = 789
)

# Develop analysis pipeline
results <- mock_data %>%
  filter(!is.na(age), !is.na(smoking)) %>%
  group_by(smoking) %>%
  summarise(
    n = n(),
    mean_age = mean(age),
    sd_age = sd(age)
  )

# Test visualisations
ggplot(mock_data, aes(x = age, fill = smoking)) +
  geom_density(alpha = 0.5) +
  labs(title = "Age distribution by smoking status (MOCK DATA)")
```

### Training new team members

Generate safe, non-sensitive data for training:

```{r}
#| eval: false

# Generate training dataset
training_data <- create_mock_data(
  databaseStart = "cchs2001_p",
  variables = "variables.csv",
  variable_details = "variable_details.csv",
  n = 500,
  seed = 111
)

# Save for training exercises
write.csv(training_data, "training_cchs_mock.csv", row.names = FALSE)
```

## Advanced features

### Specifying category proportions

Add a `proportion` column to `variable_details.csv` to control category distributions:

```csv
variable,recStart,recEnd,catLabel,proportion
smoking,1,1,Never,0.50
smoking,2,2,Former,0.30
smoking,3,3,Current,0.20
```

Without proportions, MockData generates equal probabilities for all categories.

### Survival data and custom distributions

Advanced features are specified directly in `variables.csv` using additional columns:

**Survival data parameters:**

- `event_prop`: Probability event occurs (0-1)
- `followup_min`, `followup_max`: Follow-up time range in days
- `distribution`: Distribution type (uniform, gompertz, exponential)
- `rate`, `shape`: Distribution parameters

**Example:**

```csv
uid,variable,rType,role,distribution,rate,shape,event_prop,followup_min,followup_max
ices_v02,primary_event_date,date,enabled,gompertz,0.0001,0.1,0.10,0,5475
ices_v03,death_date,date,enabled,gompertz,0.0001,0.1,0.20,365,7300
```

See [Generating survival data with competing risks](tutorial-survival-data.html) for details.

### Data quality testing

Add garbage data for testing validation pipelines using these `variables.csv` columns:

- `prop_garbage`: Simple auto-generated garbage proportion
- `garbage_low_prop`, `garbage_low_range`: Below-range invalid values
- `garbage_high_prop`, `garbage_high_range`: Above-range invalid values

See [Testing data quality and validation](tutorial-garbage-data.html) for details.

## Differences from real data

**Important limitations:**

MockData generates data that matches your metadata specifications but **does not preserve real-world statistical relationships**:

- Variables are generated independently
- No correlations between variables (e.g., age and health status)
- Joint distributions may differ from actual survey data
- Temporal patterns are simplified

**Never use mock data for:**

- Research publications
- Population inference
- Predictive modelling
- Algorithm training

**Safe uses:**

- Testing harmonisation workflows
- Developing analysis scripts
- Training team members
- Creating documentation examples

## Next steps

**Tutorials:**

- [Getting started](getting-started.html) - Learn MockData basics
- [Generating survival data with competing risks](tutorial-survival-data.html) - Time-to-event data with custom distributions
- [Working with date variables](tutorial-dates.html) - Date generation and interval notation
- [Testing data quality and validation](tutorial-garbage-data.html) - Generating garbage data for QA
- [Handling missing data](tutorial-missing-data.html) - Missing codes and proportions

**Reference:**

- [Configuration reference](reference-config.html) - Complete metadata schema documentation
- [Advanced topics](advanced-topics.html) - Derived variables, UIDs, multi-database workflows
