---
title: "Getting started with MockData v0.4"
format: html
vignette: >
  %\VignetteIndexEntry{Getting started with MockData v0.4}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| label: setup
#| include: false
input_file <- tryCatch(knitr::current_input(dir = TRUE), error = function(e) NULL)
candidate_roots <- unique(c(
  ".",
  "..",
  if (!is.null(input_file)) file.path(dirname(input_file), "..")
))
package_root <- NULL
for (candidate in candidate_roots) {
  description <- file.path(candidate, "DESCRIPTION")
  if (file.exists(description) &&
      any(grepl("^Package:\\s+MockData\\s*$", readLines(description, warn = FALSE)))) {
    package_root <- candidate
    break
  }
}

if (!is.null(package_root)) {
  devtools::load_all(package_root, quiet = TRUE)
} else {
  library(MockData)
}
```

::: {.vignette-about}
**About this vignette:** This tutorial introduces the v0.4 `mock_spec`
workflow. All code is executed when the vignette builds, so this page also
serves as a user-flow test for the public API.
:::

## The v0.4 workflow

MockData v0.4 separates data generation into three steps:

1. Create a `mock_spec`
2. Generate baseline valid values
3. Apply missing-code and garbage-value post-processing

That separation makes it easier to inspect what was requested and what was
changed after generation.

## Specify variables directly

Use the direct helpers when you want a small mock dataset without creating CSV
metadata files.

```{r}
spec <- mock_spec(
  mock_spec_continuous(
    "age",
    range = c(18, 85),
    distribution = "normal",
    mean = 50,
    sd = 12,
    rtype = "integer"
  ),
  mock_spec_categorical(
    "smoking",
    levels = c("never", "former", "current"),
    proportions = c(0.5, 0.3, 0.2),
    rtype = "character",
    missing_codes = "unknown",
    missing_proportions = 0.05
  )
)

validate_mock_spec(spec)
```

Generate baseline values first. These are values within the intended valid
space.

```{r}
baseline <- generate_mock_data_native(spec, n = 100, seed = 101)
head(baseline)
```

Then apply missing-code and garbage-value rules. This step adds diagnostics as
an attribute on the returned data frame.

```{r}
mock_data <- postprocess_mock_data(baseline, spec, seed = 102)
head(mock_data)
```

```{r}
diagnostics <- attr(mock_data, "mockdata_diagnostics")
diagnostics$variables$smoking
```

The diagnostics distinguish values assigned by post-processing from values that
were drawn naturally during baseline generation.

## Use recodeflow metadata

If you already have recodeflow-style `variables` and `variable_details`
metadata, adapt those tables to the same `mock_spec` shape.

```{r}
variables <- data.frame(
  variable = c("age", "smoking"),
  variableType = c("Continuous", "Categorical"),
  rType = c("integer", "character"),
  role = c("enabled", "enabled"),
  stringsAsFactors = FALSE
)

variable_details <- data.frame(
  variable = c("age", "smoking", "smoking", "smoking"),
  recStart = c("[18, 85]", "1", "2", "97"),
  recEnd = c("copy", "copy", "copy", "NA::b"),
  proportion = c(1, 0.6, 0.3, 0.1),
  stringsAsFactors = FALSE
)

spec_from_metadata <- mock_spec_from_recodeflow(variables, variable_details)
names(spec_from_metadata$variables)
```

The adapter preserves recodeflow semantics: valid ranges, categorical
proportions, `recEnd` missing-code rows, and garbage settings.

```{r}
metadata_baseline <- generate_mock_data_native(
  spec_from_metadata,
  n = 100,
  seed = 201
)

metadata_mock <- postprocess_mock_data(
  metadata_baseline,
  spec_from_metadata,
  seed = 202
)

head(metadata_mock)
```

```{r}
metadata_diag <- attr(metadata_mock, "mockdata_diagnostics")
metadata_diag$variables$smoking$assigned_missing_indices[1:5]
```

## Use the compatibility wrapper

`create_mock_data()` remains available for v0.3-style workflows. In strict mode,
it attempts the v0.4 pipeline for supported metadata and falls back to the legacy
`create_*` dispatch path for features that are not yet supported by the v0.4
native backend.

```{r}
wrapped <- create_mock_data(
  databaseStart = "example",
  variables = variables,
  variable_details = variable_details,
  n = 100,
  seed = 301
)

head(wrapped)
```

```{r}
!is.null(attr(wrapped, "mockdata_diagnostics"))
```

When the v0.4 path is used, `create_mock_data()` returns diagnostics. Legacy
fallback paths return plain data frames without that attribute.

## Choosing the next function

- Use `mock_*()` or `mock_spec_*()` for small examples and tests.
- Use `mock_spec_from_recodeflow()` when metadata already exists.
- Use `generate_mock_data_native()` for the default MIT-licensed backend.
- Use `generate_mock_data_simstudy()` only when the optional `simstudy` package
  is installed and you want to test that backend.
- Use `postprocess_mock_data()` when you need missing codes, garbage values, or
  diagnostics.
