---
title: "Migrate from MockData v0.3 to v0.4"
format: html
vignette: >
  %\VignetteIndexEntry{Migrate from MockData v0.3 to v0.4}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| label: setup
#| include: false
input_file <- tryCatch(knitr::current_input(dir = TRUE), error = function(e) NULL)
candidate_roots <- unique(c(
  ".",
  "..",
  if (!is.null(input_file)) file.path(dirname(input_file), "..")
))
package_root <- NULL
for (candidate in candidate_roots) {
  description <- file.path(candidate, "DESCRIPTION")
  if (file.exists(description) &&
      any(grepl("^Package:\\s+MockData\\s*$", readLines(description, warn = FALSE)))) {
    package_root <- candidate
    break
  }
}

if (!is.null(package_root)) {
  devtools::load_all(package_root, quiet = TRUE)
} else {
  library(MockData)
}
```

::: {.vignette-about}
**About this vignette:** This how-to is for users moving existing
`create_mock_data()` workflows from v0.3 to v0.4. It focuses on the compatibility
wrapper, routing messages, diagnostics, and reproducibility differences.
:::

## What stayed the same

The main entry point is still `create_mock_data()`, and the existing arguments
are still available.

```{r}
variables <- data.frame(
  variable = c("age", "smoking"),
  variableType = c("Continuous", "Categorical"),
  rType = c("integer", "character"),
  role = c("enabled", "enabled"),
  position = c(10, 20),
  distribution = c("normal", NA),
  mean = c(50, NA),
  sd = c(12, NA),
  stringsAsFactors = FALSE
)

variable_details <- data.frame(
  variable = c("age", "age", "smoking", "smoking", "smoking"),
  recStart = c("[18, 85]", "999", "1", "2", "7"),
  recEnd = c("copy", "NA::b", "copy", "copy", "NA::b"),
  proportion = c(0.95, 0.05, 0.60, 0.35, 0.05),
  stringsAsFactors = FALSE
)
```

```{r}
mock_data <- create_mock_data(
  databaseStart = "study",
  variables = variables,
  variable_details = variable_details,
  n = 100,
  seed = 123
)

head(mock_data)
```

For supported metadata, v0.4 routes this call through the new `mock_spec`
pipeline.

## See which path ran

Use `verbose = TRUE` when migrating. The message tells you whether the v0.4 path
or the legacy path was used.

```{r}
strict_data <- create_mock_data(
  databaseStart = "study",
  variables = variables,
  variable_details = variable_details,
  n = 50,
  seed = 456,
  verbose = TRUE
)
```

The v0.4 path returns a data frame with a diagnostics attribute.

```{r}
!is.null(attr(strict_data, "mockdata_diagnostics"))
```

## Opt into legacy behavior

Set `validate = FALSE` when you need the legacy v0.3 dispatch path during
migration. This is the explicit compatibility opt-out.

```{r}
legacy_data <- create_mock_data(
  databaseStart = "study",
  variables = variables,
  variable_details = variable_details,
  n = 50,
  seed = 456,
  validate = FALSE,
  verbose = TRUE
)
```

Legacy output is a plain data frame without the v0.4 diagnostics attribute.

```{r}
is.null(attr(legacy_data, "mockdata_diagnostics"))
```

The strict and legacy paths should agree on the broad shape of supported data,
but exact values can differ.

```{r}
names(strict_data)
names(legacy_data)

table(strict_data$smoking)
table(legacy_data$smoking)
```

## Understand seed differences

In v0.3, the public seed controlled the legacy generators. In v0.4, the wrapper
uses the public seed for baseline generation and `seed + 1L` for missing-code
and garbage-value post-processing.

That makes both stages reproducible, but it means exact values may differ from
v0.3 even when you pass the same seed.

```{r}
strict_again <- create_mock_data(
  databaseStart = "study",
  variables = variables,
  variable_details = variable_details,
  n = 50,
  seed = 456
)

identical(strict_data, strict_again)
```

When testing migrations, compare structure, types, ranges, and proportions
rather than expecting row-for-row equality with v0.3 output.

```{r}
str(strict_data)
prop.table(table(strict_data$smoking))
```

## Know the fallback conditions

`create_mock_data()` deliberately uses the legacy path when:

- `validate = FALSE`
- `variable_details = NULL`
- detail-level `databaseStart` filtering is needed but `variables` has no
  `databaseStart` column
- the requested metadata uses a feature not yet supported by the v0.4 native
  backend

For example, `variable_details = NULL` keeps the simple legacy fallback.

```{r}
fallback_data <- create_mock_data(
  databaseStart = "study",
  variables = variables[1, ],
  variable_details = NULL,
  n = 20,
  seed = 789,
  verbose = TRUE
)

head(fallback_data)
```

```{r}
is.null(attr(fallback_data, "mockdata_diagnostics"))
```

Unsupported v0.4 backend features also route to legacy dispatch. This example
uses an exponential continuous distribution, which remains available through the
legacy generator.

```{r}
exp_variables <- data.frame(
  variable = "time_to_visit",
  variableType = "Continuous",
  rType = "double",
  role = "enabled",
  distribution = "exponential",
  rate = 0.5,
  stringsAsFactors = FALSE
)

exp_details <- data.frame(
  variable = "time_to_visit",
  recStart = "[0, 10]",
  recEnd = "copy",
  proportion = 1,
  stringsAsFactors = FALSE
)

exp_data <- create_mock_data(
  databaseStart = "study",
  variables = exp_variables,
  variable_details = exp_details,
  n = 20,
  seed = 321,
  verbose = TRUE
)

head(exp_data)
```

## Inspect the v0.4 path directly

When debugging a migration, split the wrapper into its three v0.4 steps:

```{r}
spec <- mock_spec_from_recodeflow(variables, variable_details)
validate_mock_spec(spec, strict = TRUE)
```

```{r}
baseline <- generate_mock_data_native(spec, n = 50, seed = 456)
postprocessed <- postprocess_mock_data(baseline, spec, seed = 457)

identical(strict_data, postprocessed)
```

This makes it easier to tell whether an issue is coming from metadata parsing,
baseline generation, or post-processing.

## What to check in sibling packages

For cchsflow, chmsflow, and recodeflow workflows, test representative
`variables.csv` and `variable_details.csv` files with:

```{r}
#| eval: false
mock <- create_mock_data(
  databaseStart = "your-cycle",
  variables = "variables.csv",
  variable_details = "variable_details.csv",
  n = 100,
  seed = 123,
  validate = TRUE,
  verbose = TRUE
)

str(mock)
attr(mock, "mockdata_diagnostics")
```

Report cases where metadata unexpectedly falls back to legacy dispatch, where a
variable generated in v0.3 but errors in v0.4, or where the generated values,
types, or diagnostics are surprising.
