---
title: "Use recodeflow metadata with MockData v0.4"
format: html
vignette: >
  %\VignetteIndexEntry{Use recodeflow metadata with MockData v0.4}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| label: setup
#| include: false
input_file <- tryCatch(knitr::current_input(dir = TRUE), error = function(e) NULL)
candidate_roots <- unique(c(
  ".",
  "..",
  if (!is.null(input_file)) file.path(dirname(input_file), "..")
))
package_root <- NULL
for (candidate in candidate_roots) {
  description <- file.path(candidate, "DESCRIPTION")
  if (file.exists(description) &&
      any(grepl("^Package:\\s+MockData\\s*$", readLines(description, warn = FALSE)))) {
    package_root <- candidate
    break
  }
}

if (!is.null(package_root)) {
  devtools::load_all(package_root, quiet = TRUE)
} else {
  library(MockData)
}
```

::: {.vignette-about}
**About this vignette:** This how-to shows how to generate mock data from
recodeflow-style `variables.csv` and `variable_details.csv` files. The code
writes temporary CSV files and reads them back, so the vignette exercises the
same path as a file-based user workflow.
:::

## Starting point

Use this path when you already have recodeflow metadata. MockData reads the
metadata, converts it to a v0.4 `mock_spec`, generates baseline values, and
applies missing-code and garbage-value post-processing.

For a compact example, define three variables:

- `age`: continuous integer with a normal distribution and one missing code
- `smoking`: categorical code with one `recEnd = "NA::b"` missing-code row
- `interview_date`: date variable with a valid calendar range

```{r}
variables <- data.frame(
  variable = c("age", "smoking", "interview_date"),
  label = c("Age in years", "Smoking status", "Interview date"),
  variableType = c("Continuous", "Categorical", "Date"),
  rType = c("integer", "factor", "date"),
  role = c("enabled,table1", "enabled,table1", "enabled"),
  position = c(10, 20, 30),
  databaseStart = c("cycle1", "cycle1", "cycle1"),
  distribution = c("normal", NA, "uniform"),
  mean = c(50, NA, NA),
  sd = c(12, NA, NA),
  garbage_low_prop = c(0.02, NA, NA),
  garbage_low_range = c("[0, 17]", NA, NA),
  stringsAsFactors = FALSE
)

variable_details <- data.frame(
  variable = c(
    "age",
    "age",
    "smoking",
    "smoking",
    "smoking",
    "smoking",
    "interview_date"
  ),
  recStart = c(
    "[18, 85]",
    "999",
    "1",
    "2",
    "3",
    "7",
    "[2020-01-01, 2020-12-31]"
  ),
  recEnd = c("copy", "NA::b", "copy", "copy", "copy", "NA::b", "copy"),
  catLabel = c(
    "Valid age range",
    "Not stated",
    "Never smoker",
    "Former smoker",
    "Current smoker",
    "Don't know",
    "Interview date range"
  ),
  proportion = c(0.95, 0.05, 0.50, 0.30, 0.17, 0.03, 1),
  databaseStart = "cycle1",
  stringsAsFactors = FALSE
)
```

When `databaseStart` filtering is requested, include `databaseStart` in both
metadata tables. If the detail metadata has the filter column but the variables
metadata does not, `create_mock_data()` uses the legacy path for compatibility.

## Write metadata as CSV files

In a real project, these files already exist. Here we write them to a temporary
directory so this vignette remains self-contained.

```{r}
metadata_dir <- tempfile("mockdata-recodeflow-")
dir.create(metadata_dir)

variables_file <- file.path(metadata_dir, "variables.csv")
details_file <- file.path(metadata_dir, "variable_details.csv")

write.csv(variables, variables_file, row.names = FALSE, na = "")
write.csv(variable_details, details_file, row.names = FALSE, na = "")
```

## Inspect the normalized specification

`mock_spec_from_recodeflow()` reads either data frames or CSV file paths. It
returns a validated `mock_spec` without generating data.

```{r}
spec <- mock_spec_from_recodeflow(
  variables = variables_file,
  variable_details = details_file,
  databaseStart = "cycle1"
)

names(spec$variables)
```

The spec preserves the recodeflow pieces MockData needs: variable types,
categorical levels, proportions, valid ranges, missing-code rows, and garbage
rules.

```{r}
spec$variables$smoking$levels
spec$variables$smoking$missing_codes
spec$variables$age$range
spec$variables$age$garbage_rules
```

## Generate mock data with the compatibility wrapper

Most recodeflow users should start with `create_mock_data()`. In strict mode
(`validate = TRUE`, the default), supported metadata routes through the v0.4
pipeline.

```{r}
mock_data <- create_mock_data(
  databaseStart = "cycle1",
  variables = variables_file,
  variable_details = details_file,
  n = 200,
  seed = 123,
  verbose = TRUE
)

head(mock_data)
```

The output is a regular data frame.

```{r}
str(mock_data)
```

## Check diagnostics

When `create_mock_data()` uses the v0.4 path, the returned data frame has a
`mockdata_diagnostics` attribute. The attribute records which rows were changed
during missing-code and garbage-value post-processing.

```{r}
diagnostics <- attr(mock_data, "mockdata_diagnostics")
names(diagnostics$variables)
```

For example, `smoking` has a missing-code rule for code `7`, and `age` has a
low garbage rule.

```{r}
length(diagnostics$variables$smoking$assigned_missing_indices)
diagnostics$variables$smoking$assigned_missing_indices[1:6]

length(diagnostics$variables$age$assigned_garbage_indices$low)
diagnostics$variables$age$assigned_garbage_indices$low
```

Use the diagnostics as an audit trail, not as columns in the mock dataset. Some
base R operations and downstream tools can drop attributes, so inspect or save
diagnostics before heavy reshaping.

## Generate explicitly from the spec

The wrapper is convenient, but the v0.4 pipeline can also be called step by
step. This is useful when you want to inspect baseline values before
post-processing.

```{r}
baseline <- generate_mock_data_native(spec, n = 200, seed = 123)
head(baseline)
```

```{r}
postprocessed <- postprocess_mock_data(baseline, spec, seed = 124)
head(postprocessed)
```

The wrapper uses the same idea: the public seed controls baseline generation,
and `seed + 1L` controls post-processing.

## Database filtering

`databaseStart` filtering is exact token matching. A variable tagged for
`cycle10` will not accidentally match `cycle1`.

```{r}
variables_cycle10 <- variables
variables_cycle10$variable[1] <- "age_cycle10"
variables_cycle10$databaseStart[1] <- "cycle10"

combined_variables <- rbind(variables, variables_cycle10[1, ])
combined_details <- rbind(
  variable_details,
  transform(variable_details[variable_details$variable == "age", ],
            variable = "age_cycle10")
)

filtered_spec <- mock_spec_from_recodeflow(
  variables = combined_variables,
  variable_details = combined_details,
  databaseStart = "cycle1"
)

names(filtered_spec$variables)
```

## Troubleshooting

If `create_mock_data()` cannot use the v0.4 path, set `verbose = TRUE` to see
which path was chosen.

```{r}
#| eval: false
mock_data <- create_mock_data(
  databaseStart = "cycle1",
  variables = variables_file,
  variable_details = details_file,
  n = 200,
  seed = 123,
  verbose = TRUE
)
```

Common reasons for legacy fallback include `validate = FALSE`,
`variable_details = NULL`, detail-level `databaseStart` filtering without a
variable-level `databaseStart` column, and features that are intentionally
deferred from the v0.4 native backend.

For deeper diagnostics examples, see the diagnostics and garbage how-to when it
lands in the v0.4 documentation sprint.
