---
title: "How survival dates are generated"
format: html
vignette: >
  %\VignetteIndexEntry{How survival dates are generated}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| label: setup
#| include: false
# Load package - works in both local dev and pkgdown build
if (file.exists("../DESCRIPTION")) {
  devtools::load_all("../", quiet = TRUE)
} else {
  library(MockData)
}
```

::: {.vignette-about}
**About this article:** How MockData generates survival dates and why it works this way: where survival dates sit in the generation pipeline, how events and censoring are decided, which layer of the output a column belongs to, and what version 0.5 does not yet do. Each mechanism is shown with code that runs when the vignette is built, and hidden checks stop the build if MockData stops behaving as described. The full design record is the survival-dates decision record, `development/adr/v05-survival-dates.md` in the [MockData repository](https://github.com/Big-Life-Lab/MockData).
:::

## A survival date is derived from its anchor

A survival date depends on another generated column, its anchor. MockData therefore does not draw it with the other variables. Generation runs in four stages:

1. **Baseline:** every sampled variable, including the anchor dates.
2. **Survival:** each survival date, computed from its anchor.
3. **Formulas:** each `mockFormula` variable, which may use survival dates.
4. **Post-processing:** missing codes and garbage, applied to every column.

`create_mock_data()` runs all four. The stages can also be run one at a time:

```{r}
#| label: stages
spec <- mock_spec(
  mock_spec_date("entry", range = as.Date(c("2001-01-01", "2005-12-31"))),
  mock_spec_survival("death", anchor = "entry",
    followup_min = 0, followup_max = 3650, event_prop = 0.3),
  mock_spec_survival("event", anchor = "entry",
    followup_min = 0, followup_max = 3650, event_prop = 0.5,
    censored_by = "death"),
  mock_spec_formula("death_days", formula = "as.numeric(death - entry)")
)

baseline <- generate_mock_data_native(spec, n = 1000, seed = 1)
dated <- generate_survival_dates(baseline, spec, seed = 1)
derived <- evaluate_mock_formulas(dated, spec, seed = 1)
names(baseline)
names(dated)
names(derived)
```

```{r}
#| label: stage-claims
#| include: false
stopifnot(
  "the baseline holds only the anchor" = identical(names(baseline), "entry"),
  "the survival stage adds the survival dates" =
    setequal(setdiff(names(dated), names(baseline)), c("death", "event")),
  "the formula stage adds formula columns" =
    identical(setdiff(names(derived), names(dated)), "death_days")
)
```

Each stage adds its own columns and each variable stays one column, so survival dates need no special handling in assembly, post-processing or diagnostics. Survival dates and formulas share one dependency field, `depends_on`, and one ordering routine: a variable is computed after the variables it depends on. Each stage draws from its own random-number stream, so adding a survival date leaves the generated values of every other variable unchanged.

Post-processing is the exception. Missing codes and garbage for all columns come from one shared stream, in column order, so a survival date that carries missing codes or garbage can change which rows of later columns receive theirs:

```{r}
#| label: streams
entry <- mock_spec_date("entry", range = as.Date(c("2001-01-01", "2005-12-31")))
x <- mock_spec_continuous("x", range = c(0, 1),
  missing_codes = -99, missing_proportions = 0.3)
death_plain <- mock_spec_survival("death", anchor = "entry",
  followup_min = 0, followup_max = 10, event_prop = 0.5)
death_coded <- mock_spec_survival("death", anchor = "entry",
  followup_min = 0, followup_max = 10, event_prop = 0.5,
  missing_codes = "1900-01-01", missing_proportions = 0.3)

run_all <- function(s) {
  postprocess_mock_data(
    generate_survival_dates(generate_mock_data_native(s, n = 100, seed = 7), s, seed = 7),
    s, seed = 7
  )
}
without_survival <- mock_spec(entry, x)
generated_x_unchanged <- identical(
  generate_mock_data_native(without_survival, n = 100, seed = 7)$x,
  generate_mock_data_native(mock_spec(entry, death_coded, x), n = 100, seed = 7)$x
)
x_changed_plain <- sum(run_all(without_survival)$x != run_all(mock_spec(entry, death_plain, x))$x)
x_changed_coded <- sum(run_all(without_survival)$x != run_all(mock_spec(entry, death_coded, x))$x)
```

```{r}
#| label: stream-claims
#| include: false
stopifnot(
  "adding a survival date leaves other generated values unchanged" = generated_x_unchanged,
  "a survival date without missing codes or garbage leaves returned values unchanged" =
    x_changed_plain == 0,
  "a survival date with missing codes can shift later columns' missing codes" =
    x_changed_coded > 0
)
```

With the same seed, `x` is generated identically with or without the survival date. After post-processing, a survival date without missing codes changes `r x_changed_plain` of the 100 returned `x` values, and one with missing codes changes `r x_changed_coded`, because it draws from the shared stream before `x` does.

## How events are decided

Each survival date receives an event for exactly `floor(n * event_prop)` people, chosen at random, and is `NA` for everyone else. Follow-up times are whole days inside the window, spread by the chosen distribution (`uniform`, `exponential` or `gompertz`).

```{r}
#| label: events
n_deaths <- sum(!is.na(dated$death))
death_days <- derived$death_days[!is.na(derived$death_days)]
range(death_days)
```

```{r}
#| label: event-claims
#| include: false
stopifnot(
  "deaths are exactly floor(n * event_prop)" = n_deaths == floor(1000 * 0.3),
  "follow-up times are whole days" = all(death_days == floor(death_days)),
  "follow-up times fall inside the window" =
    all(death_days >= 0 & death_days <= 3650)
)
```

Here `r n_deaths` of 1,000 people have a death date, and every follow-up time is a whole number of days between 0 and 3,650.

A fixed count is the method of the legacy `create_wide_survival_data()`, kept in version 0.5 so that results match it. It suits mock data for testing pipelines. It cannot express a risk that differs from person to person, which a future hazard model would need.

The Gompertz parameters in the packaged minimal example (`shape = 0.1`, `rate = 0.0001`, time in days) show one limit of the ported method. The draws never exceed about three months, so every time is clamped to the start of the window ([#54](https://github.com/Big-Life-Lab/MockData/issues/54)):

```{r}
#| label: gompertz
gompertz <- mock_spec(
  mock_spec_date("entry", range = as.Date(c("2001-01-01", "2005-12-31"))),
  mock_spec_survival("death", anchor = "entry",
    followup_min = 365, followup_max = 7300, event_prop = 1,
    distribution = "gompertz", shape = 0.1, rate = 1e-4)
)
gompertz_dates <- generate_survival_dates(
  generate_mock_data_native(gompertz, n = 500, seed = 1), gompertz, seed = 1
)
gompertz_days <- unique(as.numeric(gompertz_dates$death - gompertz_dates$entry))
```

```{r}
#| label: gompertz-claims
#| include: false
# Update this claim, and the prose, when #54 is resolved.
stopifnot("#54: packaged Gompertz deaths all clamp to followup_min" =
  identical(gompertz_days, 365))
```

All 500 deaths fall `r gompertz_days` days after entry, the lower end of a window that runs to 7,300 days.

## Rules run on clean dates

After drawing a survival date, MockData applies two rules, in dependency order:

1. A date earlier than its anchor becomes `NA`.
2. If the variable has `censored_by`, it becomes `NA` wherever the censoring date is strictly earlier. A date censored by death is drawn after death, so it always compares against the final death date.

An event on the same day as the censoring date is kept:

```{r}
#| label: ties
ties <- mock_spec(
  mock_spec_date("entry", range = as.Date(c("2001-01-01", "2001-12-31"))),
  mock_spec_survival("death", anchor = "entry",
    followup_min = 20, followup_max = 20, event_prop = 1),
  mock_spec_survival("event", anchor = "entry",
    followup_min = 20, followup_max = 20, event_prop = 1,
    censored_by = "death")
)
tie_dates <- generate_survival_dates(
  generate_mock_data_native(ties, n = 100, seed = 1), ties, seed = 1
)
n_tied_events <- sum(!is.na(tie_dates$event))
```

```{r}
#| label: tie-claims
#| include: false
stopifnot("an event on the censoring date is kept" = n_tied_events == 100)
```

All `r n_tied_events` events on the same day as death are kept. The rules run before post-processing, on the dates as generated. Missing codes and garbage come afterwards, so a garbage value added to test a cleaning pipeline, such as a date before entry, stays in the output instead of being removed by the rules.

## Three layers of data

Because contamination comes last, generation produces three layers:

- **Clean truth:** the generated dates and formula columns, before contamination.
- **Observed data:** what `postprocess_mock_data()` and `create_mock_data()` return, after missing codes and garbage.
- **Analysis variables:** status, follow-up time and similar quantities that an analysis recalculates from the observed data. MockData does not produce this layer.

A formula column belongs to the clean-truth layer, even though it is returned with the observed data. Here a death date is later shown as a missing code:

```{r}
#| label: layers
layers <- mock_spec(
  mock_spec_date("entry", range = as.Date(c("2001-01-01", "2001-12-31"))),
  mock_spec_survival("death", anchor = "entry",
    followup_min = 100, followup_max = 200, event_prop = 1,
    missing_codes = "1900-01-01", missing_proportions = 0.3),
  mock_spec_formula("death_days", formula = "as.numeric(death - entry)")
)
clean <- evaluate_mock_formulas(
  generate_survival_dates(
    generate_mock_data_native(layers, n = 200, seed = 2), layers, seed = 2
  ),
  layers, seed = 2
)
observed <- postprocess_mock_data(clean, layers, seed = 2)
shown_missing <- observed$death == as.Date("1900-01-01")
recomputed_days <- as.numeric(observed$death - observed$entry)
```

```{r}
#| label: layer-claims
#| include: false
stopifnot(
  "post-processing shows some deaths as missing" = sum(shown_missing) > 0,
  "the formula column still describes the clean dates" =
    all(observed$death_days[shown_missing] >= 100 &
          observed$death_days[shown_missing] <= 200),
  "recomputing from the returned dates would give different values" =
    all(recomputed_days[shown_missing] < 0)
)
```

`r sum(shown_missing)` death dates are shown as the missing code `1900-01-01`, but `death_days` is still between 100 and 200 on all of them, because it was computed from the clean dates. Recomputed from the returned dates, it would be about `r format(round(mean(recomputed_days[shown_missing])), big.mark = ",")` days. For testing a cleaning pipeline that is what you want: the formula columns are the answer key. For analysis variables that must agree with the observed dates, compute them from the observed data after generation (`vignette("survival-dates-v05")` shows how).

## Why chained censoring is rejected

A `censored_by` target may not have `censored_by` of its own. Consider three dates, A on day 30, B on day 20 and C on day 10, with A censored by B and B censored by C. Applied in order, the rules set B to `NA`, so A is no longer compared with an earlier date and would be reported as observed on day 30, although observation ended on day 10. MockData refuses the specification instead:

```{r}
#| label: chained
chained_message <- tryCatch(
  mock_spec(
    mock_spec_date("entry", range = as.Date(c("2001-01-01", "2001-12-31"))),
    mock_spec_survival("a", anchor = "entry", followup_min = 30,
      followup_max = 30, event_prop = 1, censored_by = "b"),
    mock_spec_survival("b", anchor = "entry", followup_min = 20,
      followup_max = 20, event_prop = 1, censored_by = "c"),
    mock_spec_survival("c", anchor = "entry", followup_min = 10,
      followup_max = 10, event_prop = 1)
  ),
  error = conditionMessage
)
cat(chained_message)
```

```{r}
#| label: chained-claims
#| include: false
stopifnot("chained censoring is rejected" =
  grepl("Chained censoring is not supported", chained_message, fixed = TRUE))
```

A later version could keep every generated event time and derive observed outcomes from all censoring sources at once, which would allow several censoring sources per event.

## What version 0.5 does not do

The stage order is fixed: survival dates are computed before formulas, so a formula can use a survival date, but a survival date cannot depend on a formula. An anchor must be a sampled date:

```{r}
#| label: restriction
restriction_message <- tryCatch(
  mock_spec(
    mock_spec_date("entry", range = as.Date(c("2001-01-01", "2001-12-31"))),
    mock_spec_formula("entry_plus", formula = "entry"),
    mock_spec_survival("death", anchor = "entry_plus",
      followup_min = 0, followup_max = 10, event_prop = 1)
  ),
  error = conditionMessage
)
cat(restriction_message)
```

```{r}
#| label: restriction-claims
#| include: false
stopifnot("an anchor must be a sampled date" =
  grepl("an anchor must be a date variable", restriction_message, fixed = TRUE))
```

This matters for exposure-dependent survival, where the hazard depends on an exposure such as smoking. The design leaves room for it, but some cases need more than new parameters:

| Use case | What would need adding | Change to the design |
|---|---|---|
| Baseline smokers have a different hazard | A survival model using each person's exposure, with event occurrence following the model rather than a fixed count | Fits the existing specification and stage |
| The hazard depends on derived pack-years | Formulas computed before the survival dates that use them | Scheduling across stages |
| Someone quits smoking during follow-up | Exposure history, and a hazard that uses only the exposure known at each time | A substantial generator extension |
| Illness changes smoking, which changes later outcomes | Repeated updates to exposures and health states | A longitudinal simulation mechanism |

A causal reading of any of these would also need explicit assumptions about confounding and interventions, which the metadata does not yet express. Future survival models should keep the specification, the metadata adapters and the one-column-per-variable contract. They may need more parameters, scheduling across variable types, and separate rules for generating events and for observing them.
