---
title: "Add survival dates to your metadata"
format: html
vignette: >
  %\VignetteIndexEntry{Add survival dates to your metadata}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| label: setup
#| include: false
# Load package - works in both local dev and pkgdown build
if (file.exists("../DESCRIPTION")) {
  devtools::load_all("../", quiet = TRUE)
} else {
  library(MockData)
}
```

::: {.vignette-about}
**About this guide:** How to describe survival dates in `variables.csv`, move from `create_wide_survival_data()` to `create_mock_data()`, derive status and follow-up time, and keep the uncontaminated data. Every example runs when the vignette is built, and hidden checks stop the build if MockData stops behaving as this page describes.
:::

## Mark dates as survival dates

A survival date is a date variable whose `variables.csv` row names its entry date in the `anchor` column. It also needs a follow-up window (`followup_min`, `followup_max`, in days after the anchor) and `event_prop`, the share of people who have the event. An optional `censored_by` names a competing survival date that removes this one where it comes first.

This cohort has an entry date, a primary event that death can pre-empt, death, loss to follow-up, and administrative censoring:

```{r}
#| label: metadata
variables <- data.frame(
  variable = c("entry_date", "event_date", "death_date", "ltfu_date", "admin_censor_date"),
  variableType = "Date",
  rType = "date",
  role = "enabled",
  distribution = "uniform",
  anchor = c("", "entry_date", "entry_date", "entry_date", "entry_date"),
  censored_by = c("", "death_date", "", "", ""),
  followup_min = c(NA, 0, 0, 0, 1825),
  followup_max = c(NA, 3650, 3650, 3650, 3650),
  event_prop = c(NA, 0.4, 0.3, 0.1, 1),
  sourceFormat = "analysis",
  stringsAsFactors = FALSE
)

variable_details <- data.frame(
  variable = variables$variable,
  recStart = c("[2001-01-01,2005-12-31]", rep("[2001-01-01,2040-12-31]", 4)),
  recEnd = "copy",
  proportion = 1,
  stringsAsFactors = FALSE
)

cohort <- create_mock_data(
  databaseStart = "study",
  variables = variables,
  variable_details = variable_details,
  n = 1000,
  seed = 1
)
head(cohort)
```

```{r}
#| label: cohort-claims
#| include: false
n_deaths <- sum(!is.na(cohort$death_date))
events_after_death <- sum(cohort$death_date < cohort$event_date, na.rm = TRUE)
stopifnot(
  "all five dates are generated" =
    all(variables$variable %in% names(cohort)),
  "deaths are exactly floor(n * event_prop)" =
    n_deaths == floor(1000 * 0.3),
  "no event is recorded after a death" =
    events_after_death == 0
)
```

`r n_deaths` of 1,000 people have a death date, exactly `floor(n * event_prop)`, and `r events_after_death` events fall after a death. The `recStart` ranges on the survival dates' `variable_details` rows are not used; the follow-up window decides when events happen.

## Move from create_wide_survival_data()

Before MockData 0.5, survival dates were generated separately. The same metadata, without `anchor` and `censored_by`, went to `create_wide_survival_data()`:

```{r}
#| label: legacy
#| warning: false
legacy_variables <- variables[, setdiff(names(variables), c("anchor", "censored_by"))]

legacy <- create_wide_survival_data(
  var_entry_date = "entry_date",
  var_event_date = "event_date",
  var_death_date = "death_date",
  var_ltfu = "ltfu_date",
  var_admin_censor = "admin_censor_date",
  databaseStart = "study",
  variables = legacy_variables,
  variable_details = variable_details,
  n = 1000,
  seed = 1
)
```

To migrate, add `anchor` to every survival date's row, and `censored_by` where a competing risk applies (the legacy function always let death censor the primary event). Then call `create_mock_data()`, as in the first section.

```{r}
#| label: migration-claims
#| include: false
same_columns <- setequal(names(legacy), names(cohort))
legacy_deaths <- sum(!is.na(legacy$death_date))
stopifnot(
  "legacy and new paths produce the same columns" = same_columns,
  "both paths give floor(n * event_prop) deaths" =
    legacy_deaths == n_deaths,
  "the legacy path also removes events after death" =
    sum(legacy$death_date < legacy$event_date, na.rm = TRUE) == 0,
  "seeded values differ between the paths" =
    !identical(legacy$death_date, cohort$death_date)
)
```

Both paths produce the same five columns and the same number of deaths (`r legacy_deaths`), and both remove events that follow a death. The seeded values differ, because MockData 0.5 draws from its own random-number streams; regenerate any expected values you pinned against the old function. `create_wide_survival_data()` still works, but warns once per session that it is deprecated.

## Derive observed status and follow-up time

A death date existing is not the same as a death being observed. Observation ends at the earliest of death, loss to follow-up and administrative censoring, and a death after that point was not observed. Status and follow-up time must therefore come from the same observation window.

Formula variables compute both during generation. Each gets a `mockFormula` in `variable_details`; a formula cannot return a date, so the end of observation is written out in each:

```{r}
#| label: formula-metadata
end_of_observation <- "pmin(death_date, ltfu_date, admin_censor_date, na.rm = TRUE)"

derived_variables <- rbind(variables, data.frame(
  variable = c("death_status", "followup_days"),
  variableType = "Continuous",
  rType = c("integer", "double"),
  role = "enabled",
  distribution = NA,
  anchor = "",
  censored_by = "",
  followup_min = NA,
  followup_max = NA,
  event_prop = NA,
  sourceFormat = "analysis",
  stringsAsFactors = FALSE
))

derived_details <- rbind(
  transform(variable_details, mockFormula = ""),
  data.frame(
    variable = c("death_status", "followup_days"),
    recStart = "DerivedVar::[death_date, ltfu_date, admin_censor_date]",
    recEnd = c("Func::death_status", "Func::followup_days"),
    proportion = NA,
    mockFormula = c(
      paste0("as.integer(!is.na(death_date) & death_date == ", end_of_observation, ")"),
      paste0("as.numeric(", end_of_observation, " - entry_date)")
    ),
    stringsAsFactors = FALSE
  )
)

derived <- create_mock_data(
  databaseStart = "study",
  variables = derived_variables,
  variable_details = derived_details,
  n = 1000,
  seed = 1
)
head(derived[, c("death_date", "ltfu_date", "admin_censor_date",
                 "death_status", "followup_days")])
```

The same quantities can be computed in R after generation, which is how an analysis would compute them from observed data:

```{r}
#| label: r-derivation
end_date <- do.call(
  pmin,
  c(derived[c("death_date", "ltfu_date", "admin_censor_date")], na.rm = TRUE)
)
r_status <- as.integer(!is.na(derived$death_date) & derived$death_date == end_date)
r_days <- as.numeric(end_date - derived$entry_date)
```

```{r}
#| label: derivation-claims
#| include: false
unobserved_deaths <- sum(!is.na(derived$death_date) & derived$death_status == 0L)
stopifnot(
  "formula status matches the R derivation" =
    identical(derived$death_status, r_status),
  "formula follow-up time matches the R derivation" =
    isTRUE(all.equal(derived$followup_days, r_days)),
  "some deaths fall after censoring and are not observed" =
    unobserved_deaths > 0
)
```

The formula columns and the R calculation agree on every row. `r unobserved_deaths` of the `r sum(!is.na(derived$death_date))` death dates fall after loss to follow-up or administrative censoring, so their status is 0: counting `!is.na(death_date)` alone would have called them deaths. A death on the censoring date counts as observed.

The two approaches agree here because this metadata adds no missing codes or garbage. When it does, formula columns keep describing the generated values while the date columns are contaminated; see `vignette("survival-design-v05")`.

## Keep the uncontaminated data

`create_mock_data()` returns data after missing codes and garbage are applied. To keep the generated values as well, run the stages yourself and hold on to the data before `postprocess_mock_data()`:

```{r}
#| label: clean-truth
garbage_variables <- add_garbage(
  variables, "death_date",
  garbage_high_prop = 0.1,
  garbage_high_range = "[2090-01-01;2090-12-31]"
)
spec <- mock_spec_from_recodeflow(garbage_variables, variable_details)

clean <- generate_survival_dates(
  generate_mock_data_native(spec, n = 1000, seed = 3),
  spec, seed = 3
)
observed <- postprocess_mock_data(clean, spec, seed = 3)
```

```{r}
#| label: clean-truth-claims
#| include: false
changed_rows <- which(clean$death_date != observed$death_date)
garbage_rows <- attr(observed, "mockdata_diagnostics")$variables$death_date$assigned_garbage_indices$high
stopifnot(
  "garbage changed some death dates" = length(changed_rows) > 0,
  "only the rows diagnostics record as garbage differ" =
    setequal(changed_rows, garbage_rows)
)
```

`clean` holds the generated dates; `observed` is what `create_mock_data()` would return. They differ in `r length(changed_rows)` rows, exactly the rows the `mockdata_diagnostics` attribute records as garbage.

## Common errors

A date with survival parameters but no `anchor` stops generation with a message that names the fix:

```{r}
#| label: no-anchor-error
no_anchor_message <- tryCatch(
  create_mock_data("study", legacy_variables, variable_details, n = 10, seed = 1),
  error = conditionMessage
)
cat(no_anchor_message)
```

```{r}
#| label: error-claims
#| include: false
stopifnot(
  "a missing anchor is an error that names the fix" =
    grepl("but no anchor", no_anchor_message, fixed = TRUE),
  "the error does not suggest the legacy generator" =
    grepl("validate = FALSE is not a workaround", no_anchor_message, fixed = TRUE)
)
```

`validate = FALSE` is not a workaround: survival dates are generated only by the v0.4 pipeline, and the legacy generator stops on metadata that sets `anchor`. Survival dates and their anchors also need `sourceFormat = "analysis"` in this version.

## See also

- `vignette("tutorial-survival-data")` for a first walk through survival data
- `vignette("survival-design-v05")` for how the stage works and why
- `vignette("reference-config")` for every survival column
