---
title: "Generating survival data with competing risks"
format: html
vignette: >
  %\VignetteIndexEntry{Generating survival data with competing risks}
  %\VignetteEngine{quarto::html}
  %\VignetteEncoding{UTF-8}
---

```{r}
#| label: setup
#| include: false
# Load package - works in both local dev and pkgdown build
if (file.exists("../DESCRIPTION")) {
  devtools::load_all("../", quiet = TRUE)
} else {
  library(MockData)
}
```

::: {.vignette-about}
**About this vignette:** A first walk through survival data for cohort studies: an entry date, event dates, a competing risk (death) and censoring. You will see how survival dates are described in `variables.csv`, generate them with `create_mock_data()`, and check the rules MockData applies. All code runs during the vignette build, and hidden checks stop the build if MockData stops behaving as this page describes.
:::

## Overview

Survival analysis needs several date variables that respect each other:

- **Cohort entry date** (baseline, index date)
- **Event dates** (disease incidence, outcomes of interest)
- **Competing risks** (death prevents observation of the primary event)
- **Censoring** (loss to follow-up, administrative censoring)

`create_mock_data()` generates all of these from metadata. A survival date is a date variable whose row in `variables.csv` names, in the `anchor` column, the entry date it is generated from.

## Survival dates in the metadata

The packaged minimal example has an entry date and four survival dates:

```{r}
#| label: load-metadata
variables <- read.csv(
  system.file("extdata/minimal-example/variables.csv", package = "MockData"),
  stringsAsFactors = FALSE,
  check.names = FALSE
)
variable_details <- read.csv(
  system.file("extdata/minimal-example/variable_details.csv", package = "MockData"),
  stringsAsFactors = FALSE,
  check.names = FALSE
)

survival_columns <- c("variable", "anchor", "censored_by", "distribution",
                      "followup_min", "followup_max", "event_prop")
date_rows <- variables$rType == "date"
variables[date_rows, survival_columns]
```

Each survival date has:

- `anchor`: the entry-date variable it is generated from (`interview_date`)
- `followup_min`, `followup_max`: the follow-up window, in days after the anchor
- `event_prop`: the share of people who have the event; the rest are `NA` (censored)
- `distribution`: how follow-up times spread across the window (`uniform`, `exponential` or `gompertz`)
- `censored_by` (optional): a competing survival date. Where it comes first, this date is set to `NA`. Here `death_date` censors `primary_event_date`.

## Generating survival data

The minimal example also adds future-date garbage to its survival dates, for data-quality testing. This section removes it to show clean data; the last section puts it back.

```{r}
#| label: generate
#| warning: false
#| message: false
clean_variables <- variables
clean_variables$garbage_high_prop[date_rows] <- 0

mock <- create_mock_data(
  databaseStart = "minimal-example",
  variables = clean_variables,
  variable_details = variable_details,
  n = 2000,
  seed = 123
)

surv <- mock[, c("interview_date", "primary_event_date", "death_date",
                 "ltfu_date", "admin_censor_date")]
head(surv)
```

```{r}
#| label: generate-claims
#| include: false
death_prop <- variables$event_prop[variables$variable == "death_date"]
n_deaths <- sum(!is.na(surv$death_date))
stopifnot(
  "all five survival columns are generated" = ncol(surv) == 5,
  "deaths are exactly floor(n * event_prop)" =
    n_deaths == floor(nrow(surv) * death_prop)
)
```

Each row has an interview date (cohort entry). Survival dates that did not occur are `NA`. `death_date` has `event_prop = `r death_prop``, and `r n_deaths` of `r nrow(surv)` people have a death date: exactly `floor(n * event_prop)`, because MockData assigns a fixed number of events and shuffles who receives them.

## Competing risks and temporal rules

MockData applies two rules to each survival date:

1. **Entry is the baseline.** A date earlier than its anchor is set to `NA`.
2. **Competing risks.** If a date has `censored_by`, it is set to `NA` wherever the censoring date is earlier: death before a primary event means the event cannot be observed.

```{r}
#| label: verify-rules
events_after_entry <- all(surv$primary_event_date >= surv$interview_date, na.rm = TRUE)
deaths_after_entry <- all(surv$death_date >= surv$interview_date, na.rm = TRUE)
events_after_death <- sum(surv$death_date < surv$primary_event_date, na.rm = TRUE)
```

```{r}
#| label: rule-claims
#| include: false
death_days <- as.numeric(surv$death_date - surv$interview_date)
event_days <- as.numeric(surv$primary_event_date - surv$interview_date)
max_event_days <- max(event_days, na.rm = TRUE)
stopifnot(
  "no event before entry" = events_after_entry,
  "no death before entry" = deaths_after_entry,
  "no event recorded after a death" = events_after_death == 0,
  # #54: update these claims, and the prose below, when #54 is resolved.
  "#54: every death falls at followup_min" = all(death_days[!is.na(death_days)] == 365),
  "#54: every primary event falls within about three months" = max_event_days < 100
)
```

- All events on or after entry: `r events_after_entry`
- All deaths on or after entry: `r deaths_after_entry`
- Events recorded after a death: `r events_after_death`

In the minimal example no death comes before an event, so the competing-risk rule never applies. Its Gompertz parameters put every death exactly 365 days after entry and every primary event within `r max_event_days` days ([#54](https://github.com/Big-Life-Lab/MockData/issues/54)). This small specification shows the rule at work, with follow-up times that overlap:

```{r}
#| label: competing-risk-demo
spec <- mock_spec(
  mock_spec_date("entry", range = as.Date(c("2001-01-01", "2005-12-31"))),
  mock_spec_survival("death", anchor = "entry",
    followup_min = 0, followup_max = 3650, event_prop = 0.5),
  mock_spec_survival("event", anchor = "entry",
    followup_min = 0, followup_max = 3650, event_prop = 1,
    censored_by = "death")
)
baseline <- generate_mock_data_native(spec, n = 1000, seed = 1)
demo <- generate_survival_dates(baseline, spec, seed = 1)
n_censored <- sum(is.na(demo$event))
```

```{r}
#| label: demo-claims
#| include: false
stopifnot(
  "death removes some events" = n_censored > 0 && n_censored < 1000,
  "every removed event has an earlier death" =
    all(!is.na(demo$death[is.na(demo$event)])),
  "no remaining event falls after a death" =
    sum(demo$death < demo$event, na.rm = TRUE) == 0
)
```

Everyone would have the event (`event_prop = 1`), but `r n_censored` of 1,000 events are removed because death came first.

```{r}
#| label: admin-claims
#| include: false
n_admin_dates <- length(unique(surv$admin_censor_date))
# #55: update this claim, and the prose below, when #55 is resolved.
stopifnot("#55: administrative censoring is drawn per person" = n_admin_dates > 1)
```

One caveat about the example: `admin_censor_date` is a survival date with `event_prop = 1`, so each person gets a random date, `r n_admin_dates` distinct dates in all, rather than one study end date ([#55](https://github.com/Big-Life-Lab/MockData/issues/55)).

## Temporal violations for QA testing

With its garbage settings back, the minimal example adds future dates to its survival dates. MockData applies garbage after the temporal rules, so the violations stay in the output for your validation pipeline to find:

```{r}
#| label: temporal-violations-qa
#| warning: false
#| message: false
mock_qa <- create_mock_data(
  databaseStart = "minimal-example",
  variables = variables,
  variable_details = variable_details,
  n = 1000,
  seed = 999
)
future_threshold <- as.Date("2025-01-01")
n_future_events <- sum(mock_qa$primary_event_date > future_threshold, na.rm = TRUE)
n_future_deaths <- sum(mock_qa$death_date > future_threshold, na.rm = TRUE)
```

```{r}
#| label: qa-claims
#| include: false
stopifnot(
  "garbage adds future primary events" = n_future_events > 0,
  "garbage adds future deaths" = n_future_deaths > 0
)
```

- Future primary events: `r n_future_events`
- Future deaths: `r n_future_deaths`

The [Garbage data tutorial](tutorial-garbage-data.html#survival-data-garbage) covers survival garbage in more detail.

## Next steps

- `vignette("survival-dates-v05")`: add survival dates to your own metadata, move from `create_wide_survival_data()`, and derive observed status and follow-up time
- `vignette("survival-design-v05")`: how the survival stage works, the three layers of data, and what version 0.5 does not yet do
- `vignette("reference-config")`: every survival column
